---
title: "Comparison of student midterm and final exam scores 2014/15"
author: "Abelardo Pardo"
date: "7 December 2015"
output: html_document
---
  
```{r setup, echo = FALSE, results = 'hide', messages = FALSE, warning = FALSE}
library(ggplot2)
library(plyr)
require(effsize, quietly = TRUE)
options(digits = 2, scipen = 1)

cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  
  if (numPlots==1) {
    print(plots[[1]])
    
  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

Compare Midterm Scores for 2014 and 2015
=========================================

```{r, echo = FALSE}
data <- read.csv('../Data_files/scores.csv', header = TRUE, sep = ",")
# summary(data)

midterm_14 <- data[!is.na(data$X2014_MD),c('X2014_MD')]
final_mcq_14 <- data[!is.na(data$X2014.FMCQ),c('X2014.FMCQ')]
final_q1_14 <- data[!is.na(data$X2014.FQ1),c('X2014.FQ1')]
final_q2_14 <- data[!is.na(data$X2014.FQ21),c('X2014.FQ21')] +
  data[!is.na(data$X2014.FQ21),c('X2014.FQ21')]
final_q3_14 <- data[!is.na(data$X2014.FQ3),c('X2014.FQ3')]
final_q4_14 <- data[!is.na(data$X2014.FQ4),c('X2014.FQ4')]
#final_14 <- final_mcq_14 + final_q1_14 + final_q2_14 + final_q3_14 + final_q4_14
final_14 <- final_mcq_14 + final_q1_14 + final_q2_14 + final_q3_14 + final_q4_14

midterm_15 <- data[!is.na(data$X2015_MD),c('X2015_MD')]
final_mcq_15 <- data[!is.na(data$X2015.FMCQ),c('X2015.FMCQ')]
final_q1_15 <- data[!is.na(data$X2015.FQ1),c('X2015.FQ1')]
final_q2_15 <- data[!is.na(data$X2015.FQ2),c('X2015.FQ2')]
final_q4_15 <- data[!is.na(data$X2015.FQ4),c('X2015.FQ4')]
final_15 <- final_mcq_15 + final_q1_15 + final_q2_15 + final_q4_15

```

```{r}
#
# T.test for midterm score
# 
t.test(midterm_14, midterm_15, alternative = 'less')
wilcox.test(midterm_14, midterm_15, alternative = 'less')
cd_midterm <- cohen.d(midterm_14, midterm_15, hedges.correction = TRUE)
cd_midterm
```

FIXING!!! An independent-samples t-test was conducted to compare the rating of feedback in the two editions of the course. There was a significant difference in the rating of feedback between the answers received in the 2014 edition (M=`r mean(q8_14)`, SD=`r sd(q8_14)`) and the 2015 edition (M=`r mean(q6_15)`, SD=`r sd(q6_15)`); t(`r result$parameter`) = `r result$statistic`, p = `r result$p.value`, Cohen's d = `r cd$estimate`. These results suggest that the treatment in the second edition had a medium positive effect on how students perceived feedback.

```{r}
#
# T.test for final score
# 
t.test(final_14, final_15, alternative = 'less')
wilcox.test(final_14, final_15, alternative = 'less')
cd_final <- cohen.d(final_14, final_15, hedges.correction = TRUE)
cd_final
#
# Plot them: http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/
#

# Create single data frame with column distinguishing years
df <- data.frame(Midterm = round(midterm_14), Year = '14')
df <- rbind(df, data.frame(Midterm = round(midterm_15), Year = '15'))

cdat <- ddply(df, "Year", summarise, Midterm.mean=mean(Midterm))
p1 <- ggplot(df, aes(factor(Midterm), fill = Year)) + 
       geom_bar(aes(y = 100 * (..count..)/sum(..count..)), 
                position = "dodge",
                color = "black") +
       geom_vline(data = cdat, aes(xintercept = Midterm.mean, colour = Year),
                  linetype="dashed", size=1) +
       ggtitle("Midterm scores (out of 20)") +
       xlab("") +
       ylab("Percentage") +
       scale_colour_manual(values = cbPalette) +
       scale_fill_manual(values = cbPalette)
  
ggsave("midterm_scores.png", path = "~/Papers/2016/Presentations/Adelaide_DLW/")

#       geom_density(alpha = .3) +

df2 <- data.frame(Final = final_mcq_14 + final_q1_14 + final_q2_14 + final_q4_14,
                 Year = '14')
df2 <- rbind(df2,
            data.frame(Final = final_mcq_15 + final_q1_15 + final_q2_15 + 
                         final_q4_15, Year = '15'))

cdat2 <- ddply(df2, "Year", summarise, Final.mean=mean(Final))
p2 <- ggplot(df2, aes(x = Final, fill = Year)) + 
        geom_density(alpha = .3) +
        geom_vline(data = cdat2, aes(xintercept = Final.mean, colour = Year),
             linetype="dashed", size=1) +
        ggtitle("Final scores (out of 40)") +
        xlab("") +
        ylab("Percentage") +
        scale_colour_manual(values = cbPalette) +
        scale_fill_manual(values = cbPalette)

#multiplot(p1, p2, cols=2)
p1
p2
```
