---
title: "Comparison of student perception of feedback 2014/15"
author: "Abelardo Pardo"
date: "7 December 2015"
output: html_document
---

```{r setup, echo = FALSE, results = 'hide', messages = FALSE, warning = FALSE}
require(gdata, quietly = TRUE)
require(effsize, quietly = TRUE)

#
# Slurp the 14 data
#
df14 = read.xls("2014_detailed_data.xlsx", sheet = 1, header = TRUE)
q8_14 <- df14$Q8[1:205]    # Remove the bottom columns with the summary
q8_14 <- q8_14[is.na(q8_14) == FALSE] # Remove NA
```

Feedback 2014 (n = `r length(q8_14)`)
======================================

```{r}
hist(q8_14)
```

```{r, echo = FALSE}
#
# Slurp the 2015 data
#
df15 = read.xls("2015_1601_raw_data.xlsx", sheet = 2, header = TRUE)
q6_15 <- df15$q6[1:207]    # Remove the bottom columns with the summary
q6_15 <- q6_15[is.na(q6_15) == FALSE] # Remove NA
```

Feedback 2015 (n =  `r length(q6_15)`)
====================================

```{r}
hist(q6_15)
```

T-test: Means are equal for both variables
==========================================

```{r}
result <- t.test(q8_14, q6_15, var.equal = FALSE, paired = FALSE)
cd <- cohen.d(q6_15, q8_14, hedges.correction = TRUE)
```

An independent-samples t-test was conducted to compare the rating of feedback in the two editions of the course. There was a significant difference in the rating of feedback between the answers received in the 2014 edition (M=`r mean(q8_14)`, SD=`r sd(q8_14)`) and the 2015 edition (M=`r mean(q6_15)`, SD=`r sd(q6_15)`); t(`r result$parameter`) = `r result$statistic`, p = `r result$p.value`, Cohen's d = `r cd$estimate`. These results suggest that the treatment in the second edition had a medium positive effect on how students perceived feedback.

Additional analysis to be done. Students received personalised emails with their
progress and some suggestions on what to do with the previous activities. 
Include a sense of the type of reaction to those emails (checking the suffix in
the URLs loaded by the students). It is very possible that only a fraction of
the students did something immediate about the emails, but still, the perception
of the feedback is better (half a standard deviation).
